Report on a Safety Checker for Petri nets

### 2013-11-18 paper/workspace/main.tex
Updated main.

The method Safety checks that a given Petri net \verb=N= never violates a property \verb=P=.
We present the method Safety by example on a simplified mutual exlusion algorithm by Delzanno2001.

@@

Place m is marked in the net exactly if m is false in the algorithm.

@@

\begin{verbatim}
* Property of state constraints C: if C U {\neg P} is unsat then N |= P

@@

  s = initial number of tokens of place s - output transitions of s + input transitions of s

@@

  place      >= 0
  transition >= 0

@@

Given Petri net \verb?N? that does not violate property \verb?P?, method Invariant constructs an invariant \verb?I? with the following properties.
\begin{itemize}
\item For each reachable marking \verb=M=, \verb?I(M)? is valid.
\item For markings that violate the property, \verb?I(M)? is unsat.
\item \verb?I? is inductive:
  \verb?\forall M. I(M) valid ^ M -> M1 implies I(M1) valid?
\end{itemize}

@@

@@

  target1 = 1
  nc      = 0
  m       = 1
  cs1     = 0
  cs2     = 0

@@

* Subprocedure \mathcal{C'}:

@@

Intermediate results:

Create variable p for property P.

For t1 construct inequation `- nc - m + cs1 <= 0` in the following way.
   - input_places + output_places <= 0

For t2 construct inequation `- nc - m + cs2 <= 0` in the following way.
   - input_places + output_places <= 0

For t3 construct inequation `nc + m - cs1 <= 0` in the following way.
   - input_places + output_places <= 0

For t4 construct inequation `nc + m - cs2 <= 0` in the following way.
   - input_places + output_places <= 0

Output:
  - nc - m + cs1   <= 0
  - nc - m + cs2   <= 0
  + nc + m - cs1 <= 0
  + nc + m - cs2 <= 0

  target1 >= 0
  nc      >= 0
  m       >= 0
  cs1     >= target1
  cs2     >= target1

@@

* Theorem Dual State Constraints:

  TODO: Unroll ?
\end{verbatim}

\newpage

\iffalse
Constraints $\mathcal C'(N, P)$ with
$Y_1 = \begin{pmatrix}target1 \end{pmatrix}$,
$Y_2 =\begin{pmatrix}nc & m & cs1 & cs2\end{pmatrix}$,
$A =\begin{pmatrix}0 & 0 & 1 & 1 \end{pmatrix}$ and
$b =\begin{pmatrix}2 \end{pmatrix}$:

Invariant $\text{Inv}(N, P, A) = (cs1 + cs2 + m \le 1)$
\fi

>>>>

The method Safety checks that a given Petri net \verb=N= never violates a property \verb=P=.
We present the method Safety by example on a simplified mutual exclusion algorithm by Delzanno2001.

@@

* Petri net:

@@

\begin{verbatim}
* Property of state constraints C: If C U {\neg P} is unsat then N |= P.

@@

  # of tokens in s = initial number of tokens of place s - # times each output transition of s fires + # times each input transition of s fires

@@

  # of tokens in place s           >= 0
  # of times transition t is fired >= 0

@@

The method Invariant constructs an invariant \verb=I= for given Petri net \verb=N= and \verb=P= when \verb=N= never violates \verb=P=.
We present the method Invariant by example on a simplified mutual exclusion algorithm by Delzanno2001.

@@

\begin{verbatim}
* Property of dual state constraints C':  If C' is sat then N |= P.

* Property of invariant I:
  - I is reachable: For each reachable marking M, I(M) is valid
  - I is safe:      For markings that violate the property, I(M) is unsat
  - I is inductive: For each marking M, if I(M) is valid and M -> M1 then I(M1) is valid
\end{verbatim}

@@

  target1 = 1
  nc      = 0
  m       = 1
  cs1     = 1
  cs2     = 1

@@

\newpage
\begin{verbatim}
* Derivation of invariant I for Petri net N that satisfies property P:

  M                                    : marking          ~ places
  C                                    : incidence matrix ~ rows correspond to places, columns correspond to transitions, relates places to transitions
  X                                    : firing vector    ~ transitions

  The following constraints C1 are unsat.
  M = M0 + C*X                         : place equations
  M >= 0                               : non-negativity conditions for places
  X >= 0                               : non-negativity conditions for transitions
  AM >= b                              : property P negated

  Substitute M to obtain constraints C2.
  A(M0 + CX) >= b                      : property P negated
  M0 + CX >= 0                         : non-negativity conditions for places
  X >= 0                               : non-negativity conditions for transitions

  Rewrite each system to obtain constraints C3.
  (-A*C)      (A*M0-b)                 : property P negated
  (  -C)*X <= (  M0  )                 : non-negativity conditions for places
  (  -I)      (   0  )                 : non-negativity conditions for transitions

  Apply Farkas' Lemma to obtain constraints C4.
  yT*(-A*C)                            : 
     (  -C) = 0                        : 
     (  -I)                            :

  yT*(A*M0-b)                          : 
     (  M0  ) < 0                      : 
     (   0  )                          :

  y >= 0                               :

  The constraints C4 are sat iff the following constraints C5 are sat.
  y1 * A * C + y2 * C + y3 = 0         : 
  y1 * (A*M - b) + y2 * M0 < 0         : 
  y1 >= 0                              : 
  y2 >= 0                              : 
  y3 >= 0                              :

  The constraints C5 are sat iff the following constraints C6 are sat.
  (y1 * A + y2) * C  <= 0              : 
  (y1 * A + y2) * M0 < y1 * b          : 
  y1 >= 0                              : 
  y2 >= 0                              :

  The constraints C6 are sat iff the following constraints C' are sat.
  \lambda * C  <= 0                    : inductivity constraint
  \lambda * M0 < y1 * b                : safety constraint
  \lambda >= y1 * A                    : property constraint
  y1 >= 0                              : non-negativity constraint

  For \lambda satisfying C' the invariant is the following.
  I(M) = (\lambda * M <= \lambda * M0) : invariant
\end{verbatim}

\newpage
\begin{verbatim}
* Subprocedure \mathcal{C'}:

@@

TODO: Write the code

@@

  Input:
    (S, T, E, I) : Petri net
    A            : Satisfying assignment for C'

  Output:
    I            : Invariant

TODO: Write the code
\end{verbatim}

\newpage
\section{Method Safety by Refinement}

The method Safety by Refinement applies trap conditions to check that a given Petri net \verb=N= never violates a property \verb=P=.
We present the method Invariant by example on a mutual exclusion algorithm by Delzanno2001.

\begin{verbatim}

* Code

Property: Thread 1 and Thread 2 are never in any critical section at the same time.

* Petri net:
\end{verbatim}
\begin{center}
\begin{tikzpicture}[node distance=2cm]
  \node[blue place, tokens=2, label=above:nc] (nc) {};
  \node[place, tokens=1, label=right:m2, below of=nc, xshift=+0.75cm] (s2) {};
  \node[place, tokens=1, label=left:m1, below of=nc, xshift=-0.75cm] (s1) {};
  \node[red place, label=above:cs1, left of=s1] (cs1) {};
  \node[red place, label=above:cs2, right of=s2] (cs2) {};
  \node[transition, label=above:t1] (t1) at ($(s1)!0.5!(cs1) + (0,1cm)$) {}
	edge [pre] (nc)
	edge [pre] (s1)
	edge [pre] (s2)
	edge [post] (cs1)
	edge [post] (s2);
  \node[transition, label=above:t2] (t2) at ($(s2)!0.5!(cs2) + (0,1cm)$) {}
	edge [pre] (nc)
	edge [pre] (s2)
	edge [pre] (s1)
	edge [post] (cs2)
	edge [post] (s1);
  \node[transition, label=below:t3] (t3) at ($(s1)!0.5!(cs1) - (0,1cm)$) {}
	edge [pre] (cs1)
	edge [post] (s1)
	edge [post,bend left,min distance=3cm, out=115, in=115] (nc);
  \node[transition, label=below:t4] (t4) at ($(s2)!0.5!(cs2) - (0,1cm)$) {}
	edge [pre] (cs2)
	edge [post] (s2)
	edge [post,bend right,min distance=3cm, out=-115, in=-115] (nc);
\end{tikzpicture}
\end{center}

  Subprocedure TrapConditions constructs trap conditions C_\theta corresponding to N and A.
  Subprocedure \Delta constructs constraint \delta corresponding to A_\theta.
\end{verbatim}
\begin{center}
\begin{tikzpicture}
  \node[state] (begin) {BEGIN};
  \node[action, below=of begin] (c) {$C:=\mathcal C(N)$};
  \node[decision, below=of c] (satc) {$\text{SAT}(C \cup \{\neg P\})$};
  \node[action, below=of satc] (modelc) {$A:=\text{Model}(C \cup \{\neg P\})$\\
                               	$C_{\theta}:=\text{TrapConditions}(N, A)$};
  \node[decision, below=of modelc] (satctheta) {$\text{SAT}(C_\theta)$};
  \node[action, below=of satctheta] (modelctheta) {$A_\theta:=\text{Model}(C_\theta)$\\
                                    	$\delta:=\Delta(A_\theta)$\\
                                    	$C:=C \cup \{\delta\}$};
  \node[print, right=of satc] (yes) {YES};
  \node[print, right=of satctheta] (dontknow) {Don't know};
  \node[state, right=of yes] (end1) {END};
  \node[state, right=of dontknow] (end2) {END};

  \draw (begin) edge (c);
  \draw (c) edge coordinate[pos=.5] (edgein) (satc);
  \draw (satc) edge node[above]{NO} (yes);
  \draw (yes) edge (end1);
  \draw (satc) edge node[right]{YES} (modelc);
  \draw (modelc) edge (satctheta);
  \draw (satctheta) edge node[above]{NO} (dontknow);
  \draw (dontknow) edge (end2);
  \draw (satctheta) edge node[right]{YES} (modelctheta);
  \draw (modelctheta.south) -- ([yshift=-0.5cm] modelctheta.south)
  -| ([xshift=-2cm] modelctheta.west) |- (edgein);
\end{tikzpicture}
\end{center}

\begin{verbatim}
* Property of trap conditions C_\theta: If C_\theta is sat then there is a set S such that
  1. S is a trap in the net N
  2. S is marked in the initial marking M0
  3. S is unmarked in the assignment A

* Property of constraint \delta: Constraint \delta refines the abstraction, i.e.
  1. A ^ \delta is unsat (\delta excludes A)
  2. N |= \delta is sat  (\delta is a property of N)
\end{verbatim}

\begin{verbatim}
* State constraints C:

  nc  =  2 - t1 - t2 + t3 + t4
  m1  =  1 - t1 + t3
  m2  =  1 - t2 + t4
  cs1 =  0 + t1 - t3
  cs2 =  0 + t2 - t4

  nc  >= 0
  m1  >= 0
  m2  >= 0
  cs1 >= 0
  cs2 >= 0
  t1  >= 0
  t2  >= 0
  t3  >= 0
  t4  >= 0

* Assignment A:
  nc  = 1
  m1  = 0
  m2  = 0
  cs1 = 1
  cs2 = 1
  t1  = 1
  t2  = 1
  t3  = 0
  t4  = 0

* Trap conditions C_\theta:
 TODO

* Assignment A_\theta:
 TODO

* Constraint \delta:
 TODO
\end{verbatim}
